Linear response theory for core set identification

### 2012-04-26 writedown/linear.response/main.tex
init text

\documentclass[aps,pre,preprint,unsortedaddress]{revtex4}
% \documentclass[aps,pre,twocolumn]{revtex4-1}
% \documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[dvips]{graphicx}
\usepackage{color}
\usepackage{tabularx}
\usepackage{algorithm}
\usepackage{algorithmic}

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{blue} #1}}
\renewcommand{\v}[1]{\textbf{\textit{#1}}}
\renewcommand{\d}[1]{\textsf{#1}}

@@

\title{Onsager regression method via Adaptive Resolution Simulation}

\begin{abstract}
We combine the Onsager regression method with the adaptive resolution simulation scheme.
\end{abstract}

@@

\section{Theoretical Principles}
Let us suppose we have a system which is in equilibrium with a reservoir of energy and particles (i.e. Grand Canonical ensemble), at a certain time we apply a perturbation to the system. Such a perturbation will induce a response from the system. Let us identify $\hat{O}({\bf r,p})$ as the physical observable which expresses the response to the perturbation. If the system was characterized by a constant number of particle $\hat{N}$, then the average response of the system at time $t$ would be expressed by the equation:
\begin{equation}
O(t)=\int d{\bf r}d{\bf p} \hat{O}({\bf r,p})\rho_{\hat{N}}({\bf r,p};t)
\label{o1}
\end{equation}

@@

\begin{equation}
\int d{\bf r}d{\bf p} \hat{O}({\bf r,p})\rho_{\hat{N}}({\bf r,p};t)=\int d{\bf r}d{\bf p} \hat{O}({\bf r}(t),{\bf p}(t))\rho_{\hat{N}}({\bf r,p};0)
\label{o2}
\end{equation}
with:
\begin{equation}
\rho_{\hat{N}}({\bf r,p};0)=\rho_{\hat{N}}({\bf r,p}) lim_{T \to\infty}\frac{1}{T}\int_{0}^{T} d\tau \hat{O}\left({\bf r}(t; {\bf r}_{0}(\tau), {\bf p}_{0}(\tau)),{\bf p}(t;{\bf r}_{0}(\tau), {\bf p}_{0}(\tau)\right)
\label{rhon}
\end{equation} 
with $\rho_{\hat{N}}({\bf r,p})$ the standard canonical distribution at fixed $\hat{N}$. However in the present case, already before applying a perturbation we have a system whose total number of particles $N$ fluctuates, moreover when we apply the pertubation to such system the average response needs to be sampled over all possible realization (corresponding to each value of $N$), thus the formulas above are likely to be modified as:
\begin{equation}
O(t)=\left<\left<\hat{O}_{N}(t)|N\right>\right>_{N}=\sum_{N=0}^{\infty}P(N)\int d{\bf r}d{\bf p}\rho_{N}({\bf r,p})\hat{O}({\bf r}(t),{\bf p}(t))
\label{newvers}
\end{equation}
where:
\begin{equation}
P(N)=\frac{e^{\beta\mu N}Q_{N}}{Q}
\end{equation}
and
\begin{equation}
Q=\sum_{N}e^{\beta\mu N}Q_{N}
\end{equation}
with
\begin{equation}
Q_{N}=\int d{\bf r}d{\bf p} \frac{e^{-\beta H_{N}}}{N!h^{3N}}
\end{equation}
and
\begin{equation}
\rho_{N}({\bf r,p})=\frac{e^{-\beta H_{N}}}{N}.
\end{equation}
The essential meaning of Eq.\ref{newvers} is that we let the system evolve and calculate the average, counting the contribution of each $\hat{O}$ at each realization of $N$, weighted by the corresponding distribution for this realization $N$. So far this is the standard average in a Grand Canonical ensemble. The formulation above is nothing else than the extension of the Onsager-Kubo equation to the Grand Canonical case.
Let us discuss this in detail. Since our central quantities are $\rho_{N}$ , and in general all operators are of the form $X_{N}$ ($X$ generic operator), let us apply the procedure behind the Onsager-Kubo equation to quantities that are $N$ labeled, meaning that they depend on $N$. Liouville equation can then be written as:
\begin{equation}
\frac{\partial\rho_{N}}{\partial t}=i L_{N}\rho_{N}=iL_{N}^{0}\rho_{N}+iL_{N}^{P}\rho_{N}=\{H_{N}^{0},\rho_{N}\}+\{H^{P}_{N},\rho_{N}\}
\label{liouville}
\end{equation}
where $0$ and $p$ are indicating the unperturbed $L$ or $H$ operator and the perturbed one respectively.
For the distribution of the system we have:
\begin{equation}
\rho_{N}({\bf r},{\bf p},t)=S^{\dagger}_{N}(t)\rho_{N}({\bf r},{\bf p},0)
\end{equation}
% \bluec{ Noted by Han: This may be problematic, because I am not sure
%   if the probabilty density of an OPEN subsystem can be evolved by the
%   Liouville operator for a subsystem with FIXED number of particles
%   $N$.  Actually this equation assumes the number of particle in the
%   system is a constant, which may not be the case under perturbation.}
An observable $\hat{O}$ of the system evolves as:
\begin{equation}
\hat{O}_{N}(t)=\hat{O}({\bf r}(t),{\bf p}(t),N)=S_{N}(t)\hat{O}(0)
\end{equation}
The non equilibrium average of a given observable, $N$-labeled, i.e. $N$-dependent is obtained via:
\begin{equation}
\frac{d \hat{J}_{N}}{dt}=iL_{N}\hat{J}_{N};\hat{J}_{N}(t)=S_{N}(t)\hat{J}_{N}
\end{equation}
this could be calculated via an "equilibrium average as (Onsager-Kubo equation):
\begin{equation}
\hat{J}_{N}(t)=<\hat{J}>_{N}
\end{equation}
that is
\begin{equation}
<\hat{J}>_{N}=\int d\Gamma_{N}\hat{J}_{N}\rho_{N}(t)=(\hat{J}_{N},S^{\dagger}_{N}(t)\rho_{N}^{0})
\end{equation}
where $\Gamma_{N}$ is the phase space of the system, and finally:
\begin{equation}
 (\hat{J}_{N},S^{\dagger}_{N}(t)\rho_{N}^{0})=(S_{N}(t)\hat{J}_{N}, \rho_{N}^{0}).
\end{equation}
Now its average over the realizations $N$ follows as:
\begin{equation}
\left<\left<\hat{J}(t|N)\right>\right>_{N}=\sum_{N}p(N)\hat{J}_{N}(t)=\sum_{N}\int d{\bf r}d{\bf p}\hat{J}_{N}(t)\rho_{N}({\bf r},{\bf p}, t=0)
\end{equation}
where it must be noticed that $\rho_{N}({\bf r},{\bf p}, t=0)$ is the $\rho_{equil}({\bf r},{\bf p})$.
\bluec{Noted by Han:
  It is unclear for me that maybe $p(N)$ is also a function of time
  under perturbation. How to deal with it?}
This obvious extension is physically correct.
Let us see how to implement it. We have taken an initial condition with a given number of particles. Then, starting from this initial condition we allowed the system to freely evolve (with no constraints on $N$) and let the system go to the Grand Canonical equilibrium. Then we picked from time to time at regular intervals of time, initial conditions for the perturbed dynamics each with its $N$ kept fixed. After certain time we will have picked up a statistically significant realizations with different possible $N$ from the distribution $P(N)$.\\
The difference between the Grand Canonical approach and the Canonical approach is that:\\
(a)We need long non equilibrium trajectories to be sure that we sample enough segments with given $N$ along the non equilibrium trajectories\\
(b)A number of initial conditions with different $N$ large enough to be meaningful, that is, a dense branching along the Grand Canonical equilibrium trajectory.\\
In any case the advantage of this approach compared to an all atom simulation is the drastic reduction of computational cost of each simulation.

\section{An alternative consideration for the explicit (atomistic) reference system}
If we assume that the Observable we are interested can be expend by:
\begin{align}
  \hat O(\v r, \v p) =
  \sum_i\hat O^{(1)}(\v r_i, \v p_i) +
  \sum_{i\neq j}\hat O^{(2)}(\v r_i, \v p_i; \v r_j, \v p_j),
\end{align}
namely, we can explicitly writen down the single and pairwise
component of the observable (this is the case for most of the
properties, such as energy, pressure, density, RDF, etc.). We further
assume $\hat O^{(1)}$ and $\hat O^{(2)}$ are smooth up to the infinite
order (this may not be the case, but it is possible to polish these
observables and consider the limit of the parameter in the end. So the
smoothness of the observables is not essential in this argument).

Now assume that the grand-canonical system as a subsystem of an
``infinitely'' large canonical system. We denote the domain of the
subsystem by $\Omega$. $\v r_1, \cdots, \v r_N, \v p_1, \cdots, \v
p_N$ are all degrees of freedoms in the large canonical system. In the fully atomistic system, the
probability density is govened by a Liouville equation.  The
observable of the interested sub grand canonical can be written as a
function of the degrees of freedoms of the large canonical system,
revised by some proper screening function that shields the unconcerned
degrees of freedoms outside of the subsystem:
\begin{align}
  \hat O_{\textrm{sub-gc}}(\v r, \v p) =
  \sum_i
  \chi_\Omega(\v r_i)\,\hat O^{(1)}(\v r_i, \v p_i) +
  \sum_{i\neq j}
  \frac12 \,[\,
  \chi_\Omega(\v r_i) + \chi_\Omega(\v r_j)
  \,]\,
  \hat O^{(2)}(\v r_i, \v p_i; \v r_j, \v p_j),  
\end{align}
where $\chi_\Omega$ is defined by
\begin{align}
  \chi_\Omega(\v r)=
  \left\{
    \begin{array}{lcl}
      1 &\qquad\qquad& \v r \in \Omega; \\
      0 && \textrm{otherwise}.
    \end{array}
  \right.
\end{align}
Again, though the screening funciton $\chi_\Omega$ is not smooth, it
can be polished by some infinitly smooth function and consider the
limit of the polishing parameter in the end.  Following the argument
for the nonequilibrium canonical system:
% , we have the similar argument
% for the grand canonical subsystem:
\begin{align}\nonumber
  O_{\textrm{sub-gc}}(t)
  & = \langle \hat O_{\textrm{sub-gc}}(\v r, \v p)\rangle_t\\\nonumber
  & =
  \int
  \hat O_{\textrm{sub-gc}}(\v r, \v p) \rho(\v r, \v p, t)\,
  \d d\v r\d d\v p\\\nonumber
  & =
  \int
  \hat O_{\textrm{sub-gc}}(\v r, \v p) S(t)\rho(\v r, \v p)\,
  \d d\v r\d d\v p\\\nonumber
  & =
  \int
  S^{\dagger}(t)\hat O_{\textrm{sub-gc}}(\v r, \v p) \rho(\v r, \v p)\,
  \d d\v r\d d\v p\\
  & =
  \int
  \hat O_{\textrm{sub-gc}}(\v r(t), \v p(t)) \rho(\v r, \v p)\,
  \d d\v r\d d\v p
\end{align}
The $\rho$ is the standard canonical density distribution, which
includes all information of the grand canonical subsystem (especially
$p(N)$). The point is that we are able to study the observable of the
grand canonical subsystem (as a function of time), by the
non-equilibrium time evolution of the large canonical system, whose
governing function is well defined (do not need to consider the
exchange of particles, which is considered by $\rho(\v r, \v p, t)$
automatically).

\bibliography{ref}{}
\bibliographystyle{unsrt}

>>>>

% \documentclass[aip,jcp,preprint,unsortedaddress,a4paper,onecolum]{revtex4-1}
\documentclass[aip,jcp,a4paper,reprint,onecolumn]{revtex4-1}
% \documentclass[aps,pre,twocolumn]{revtex4-1}
% \documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage[dvips]{graphicx}
\usepackage{color}
\usepackage{tabularx}
\usepackage{algorithm}
\usepackage{algorithmic}

\makeatletter
\makeatother

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{blue} #1}}
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dd}{\textsf{d}}

\newcommand{\mh}{\mathcal H}
\newcommand{\ml}{\mathcal L}
\newcommand{\mt}{\mathcal T}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}

@@

\title{Linear response theory for core set identification}
\author{Han Wang}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Berlin, Germany}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Berlin, Germany}

\begin{abstract}
  % In this draft, we are trying to study the core set based Markov
  % State Model (MSM) under non-equilirbium conditions.
\end{abstract}

@@

\section{Theoretical considerations}
\subsection{The linear response theory}
In this sub-section, we give a brief desciption of the linear response
theory, for details see Ref.~\cite{tuckeman2010statistical} for
example. Here we assume ths system is goverened by equation:
\begin{align}\label{eqn:gov-1}
  \dot{\vect q}_i & = \frac{\partial\mh}{\partial \vect p_i}
  + \vect C_i(\vect q,\vect p) F_e(t)\\\label{eqn:gov-2}
  \dot{\vect p}_i & =- \frac{\partial\mh}{\partial \vect q_i}
  + \vect D_i(\vect q,\vect p) F_e(t)
\end{align}
Where $\vect C$ and $\vect D$ are perturbations to the system, and
$F_e(t)$ is the strength of the perturbation as a function of time.
The phase space imcompressibility condition is assumed:
\begin{align}
  \sum_{i=1}^{3N}
  \bigg[
  \frac{\partial \vect C_i}{\partial \vect q_i} +
  \frac{\partial \vect D_i}{\partial \vect p_i}
  \bigg] = 0
\end{align}
Then the phase space distribution $f(\vect x, t)$ ($\vect x = \{\vect
q_i, \vect p_i\}$) is subject to the Liouville equation:
\begin{align}
  \frac{\partial}{\partial t} f(\vect x, t) + i\ml f(\vect x, t) = 0
\end{align}
where $\ml$ is the Liouville operator given by $i\ml = \dot{\vect
  x}\cdot\nabla_{\vect x}$.  We assume that the perturbation to the
system is so small that both the phase space distribution and the
Liouville operator can be viewed as perturbation from the quilibirum
distribution (invariant measure) and the unperturbed operator,
respectively. Namely
\begin{align}
  f(\vect x, t) &= f_0(\vect x) + \Delta  f(\vect x, t)
\end{align}
and since the Eqn.~\eqref{eqn:gov-1} and \eqref{eqn:gov-2} imply that
$\dot{\vect x} = \dot{\vect x}_0 + \Delta \dot{\vect x}(t)$
\begin{align}
  i\ml =
  (\dot{\vect x}_0 + \Delta \dot{\vect x}(t))\cdot\nabla_{\vect x}
  = i\ml_0 + i\Delta\ml(t)
\end{align}
Where the invariant measure satisfies $i\ml_0f_0(\vect x) = 0$. Also
$f_0(\vect x)$ has the form $f_0(\mh(\vect x))$ (?). The Liouville equation
for the perturbed system is therefore:
\begin{align}
  \frac{\partial}{\partial t}
  [f_0(\mh(\vect x)) + \Delta  f(\vect x, t)]
  +
  [i\ml_0 + i\Delta\ml(t)]
  [f_0(\mh(\vect x)) + \Delta  f(\vect x, t)] = 0
\end{align}
The basic assumption for the ``linear response'' is that the high order
term $i\Delta\ml(t)  \Delta  f(\vect x, t)$ in the Liouville equation
can be neglected. Noticing $i\ml_0 f_0 = 0$, we have the Liouville equation
for the leading order of perturbation:
\begin{align}
  \bigg[
  \frac{\partial}{\partial t}
  + i\ml_0
  \bigg]
  \Delta  f(\vect x, t)
  =
  - i\Delta\ml(t) f_0(\mh(\vect x))
\end{align}
The operator on l.h.s. is the same as the unperturbed Liouville equation.
Formally solving this equation with initial condition
$\Delta f(\vect x, 0) = 0$ gives
\begin{align}
  \Delta  f(\vect x, t) =
  - \int_0^t \dd s \:
  e^{-i\ml_0(t-s)}
  i\Delta\ml(s)f_0(\mh(\vect x))
\end{align}
By defining the \emph{dissipative flux},
\begin{align}
  j(\vect x) =
  -\sum_{i=1}^{3N}
  \bigg[
  \vect D_i(\vect x) \frac{\partial\mh}{\partial \vect p_i} +
  \vect C_i(\vect x) \frac{\partial\mh}{\partial \vect q_i}   
  \bigg]
\end{align}
it has been shown that~\cite{tuckeman2010statistical}
\begin{align}
  i\Delta\ml(s)f_0(\mh(\vect x)) =
  -\frac{f_0(\mh)}{\partial\mh} j(\vect x)\,F_e(s)
\end{align}
For the canonical invariant measure, this is equivalent to
\begin{align}
  i\Delta\ml(s)f_0(\mh(\vect x)) =
  \beta f_0(\mh(\vect x)) j(\vect x)F_e(s)
\end{align}
For any observation:
\begin{align}\nonumber
  \mathcal A(t)
  &=
  \int\dd \vect x A(\vect x)f(\vect x, t)  \\\nonumber
  &=
  \int\dd \vect x\, A(\vect x)\,
  [f_0(\mh(\vect x)) + \Delta f(\vect x, t)] \\\nonumber
  &=
  \langle A\rangle_0
  -
  \beta
  \int \dd \vect x
  A(\vect x)
  \int_0^t\dd s\,
  e^{-i\ml_0(t-s)}
  f_0(\mh(\vect x)) j(\vect x)F_e(s)  \\\nonumber
  &=
  \langle A\rangle_0
  -
  \beta
  \int_0^t\dd s\,
  F_e(s)
  \int \dd \vect x\,
  A(\vect x)
  e^{-i\ml_0(t-s)}
  f_0(\mh(\vect x)) j(\vect x)  \\
  &=
  \langle A\rangle_0
  -
  \beta
  \int_0^t\dd s\,
  F_e(s)
  \int \dd \vect x\,
  f_0(\mh(\vect x))
  A(\vect x)
  e^{-i\ml_0(t-s)}
  j(\vect x)  
\end{align}
Notice that $e^{i\ml_0(t-s)}A(\vect x) = A(\phi_{t-s}(\vect x))$,
where $\phi_t(\vect x)$ is the flow function of the \emph{unperturbed}
Hamiltonian system. We have
\begin{align}
  \mathcal A(t)
  =
  \langle A\rangle_0
  -
  \beta
  \int_0^t\dd s\,
  F_e(s)
  \int \dd \vect x\,
  f_0(\mh(\vect x))
  A(\phi_{t-s}(\vect x))
  j(\vect x)  
\end{align}

@@

\begin{align}
  \mathcal A(t)
  =
  \langle A\rangle_0
  -
  \beta
  \int_0^t\dd s\,
  F_e(s)
  \langle
  A(t-s)\,
  j(0)\,
  \rangle_0
\end{align}
where $\langle A(t-s)\,j(0)\,\rangle_0$ is the \emph{equilibrium time
  correlation} between the observation $A$ at time $t-s$ and the
dissipative flux $j$ at time 0.

\subsection{The Core set as an observation of the system}

The core set plays a very important role in the MSM theory. One way to
identify core set is to consider the ``stability'' of the invariant
measure $f_0$ under some mapping $\mt_\alpha$ of time step $\alpha$:
\begin{align}
  C_\alpha = \big\{
  \vect x \,\vert\, \mt_\alpha f_0(\vect x) > f_0(\vect x)
  \big\}
\end{align}
Generally, it is impossible and not necessary to consider the core set
in the $3N$ dimensional configuration space. For example, a
macromolecule solving in water, the DOFs of water is not of interest
at all. Therefore, we consider some phase space compressing function
$\vect y = \vect y(\vect x)$, with $\vect y \in \Omega \subset \mathbb
R^M,\ M\ll 3N$. We call $\Omega$ the observation space.  The projected
invariant measure on the observation space is therefore
\begin{align}
  \mathcal P f_0 (\vect y)
  =
  \int
  \dd \vect x\,
  f_0(\vect x) \,
  \delta (\vect y - \vect y(\vect x)) 
\end{align}
The core set on the observation space is
\begin{align}
  \widetilde
  C_\alpha = \big\{
  \vect y \,\vert\,
  \mathcal P\mt_\alpha f_0(\vect y) > \mathcal P f_0(\vect y)
  \big\}
\end{align}
% with $\widetilde \mt_\alpha$ defined by
% $\widetilde\mt_\alpha = \mathcal P\circ\mt_\alpha$, namely
% \begin{align}
%   \widetilde\mt_\alpha\, p_0(\vect y)
%   =
%   \mathcal P\circ\mt_\alpha f_0 (\vect y)
%   =
%   \int
%   \dd \vect x\,
%   [\mt_\alpha f_0](\vect x) \,
%   \delta (\vect y - \vect y(\vect x)) 
% \end{align}
For convenience, we define $\Delta \mt_\alpha =
\mt_\alpha - \mathcal I$, where $\mathcal I$ is the
identity. If $[\mathcal P\circ\Delta \mt_\alpha f_0](\vect y) > 0$, then
$\vect y$ is in the core set $\widetilde C_\alpha $.
\begin{align}\nonumber
  [\mathcal P\circ\Delta \mt_\alpha f_0](\vect y)
  &=
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  \delta (\vect y - \vect y(\vect x))  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  K_{\sigma} (\vect y - \vect y(\vect x))  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  K_{\sigma, \vect y} (\vect x)  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  f_0(\vect x) \,
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y} (\vect x)\\
  &=
  \lim_{\sigma \rightarrow 0}\,
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_0
\end{align}
where $[\Delta \mt_\alpha]^\dagger$ is the adjoint operator of $\Delta
\mt_\alpha$, which is actually $\Delta \mt_{-\alpha}$. The delta
function is replaced by the limit of a mollifier $K_\sigma(\vect y)$,
which is inifinely smooth, compact supported, bounded and $\int
K_\sigma(\vect y)\dd \vect y = 1$.  The physical meaning is that we
can consider the expectation value of the stability of the molifier
$K_{\sigma, \vect y}$ instead of the stability of the invariant measure.
This provides the covenience of using the linear response theory showed
in the previous section:
\begin{align}\nonumber
  [\mathcal P\circ\Delta \mt_\alpha f_t](\vect y)
  &=
  \lim_{\sigma \rightarrow 0}\,
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_t \\
  &=
  \lim_{\sigma \rightarrow 0}\,
  \Big\{
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_0  -
  \beta
  \int_0^t\dd s\,
  F_e(s)
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}(t-s)\,
  j(0)\,
  \rangle_0
  \Big\}
\end{align}

\section*{References}
\bibliography{ref}{}
\bibliographystyle{unsrt}
